### 2024-06-26 parts/intro.tex
update intro

\AP There have been recurrent attempts to generalize the notions of regularity
from languages to functions, leading to classes of various expressiveness, such
as\footnote{This list is far from exhaustive.} Mealy Machines ($\intro*\Mealy$)
\cite{MEAL55}, sequential functions ($\intro*\Sequential$) \cite{SCHU77},
rational functions ($\intro*\Rational$) \cite{EILE74}, \intro{regular functions}
($\intro*\Regular$) \cite{ENMA02}, and \intro{polyregular functions} ($\intro*\Poly$)
\cite{BOKL19}. For all of these models, an \emph{aperiodic} counterpart can be
defined (respectively, $\intro*\AMealy$, $\intro*\ASequential$,
$\intro*\ARational$, $\intro*\ARegular$, $\intro*\SF$), lifting the
correspondence between counter-free automata, star-free languages and aperiodic
monoids to the functional setting
\cite{FKT14,BOJA14,CADA15,DJR16,BDK18,BOKL19,DGK21}. The inclusions between
these classes of functions are all known to be strict, and we depicted in
\cref{computational-models:fig} the status of the \emph{membership problem}
associated to these strict inclusions (that is, given a function $f$, decide if
$f$ can be computed by a function of a proper subclass). Note that in
\cref{computational-models:fig}, if $\mathsf{B}$ is a proper subclass of
$\mathsf{C}$, then the \emph{aperiodic} variant of $\mathsf{B}$ is precisely
the intersection of $\mathsf{B}$ with the \emph{aperiodic} variant of the
larger class $\mathsf{C}$. As an example, to conclude that a function can be
computed by an \emph{aperiodic sequential function} ($\ASequential$) it is
enough to prove that it is sequential ($\Sequential$), and computed by an
\intro{aperiodic polyregular function} ($\SF$).

\AP To approach the \emph{aperiodicity} membership problems the traditional
approach has been to develop so-called canonical objects, akin to the minimal
automaton or the syntactic monoid of regular languages, over which aperiodicity
is equivalent to a simple syntactic criterion (e.g. having no counters). This
is the proof scheme that has been implemented for sequential functions
\cite{CHOF03}, and more recently for rational functions \cite{FGL16,FGLM18}.
Unfortunately, such a canonical object remains to be defined for regular and
polyregular functions.

@@

\AP In order to approach the decision problems on
\kl{polyregular functions}, restricting the output alphabet to a single letter
has proven to be a fruitful approach \cite{DOUE21,DOUE22}. The key advantage of
single-letter output alphabets is that the particular order in which letters
are produced by the machine is irrelevant. This allows to obtain results that
would otherwise fail because of this ordering, such as the so-called
\emph{pebble minimization theorem} of \cite{DOUE21}, which fails in the general
case \cite{BOJA22,KLEP23}. Let us remark that the restriction to \emph{unary
output} in the case of \kl{polyregular functions} can often be lifted to
outputs in the free commutative semigroup generated by a finite alphabet
$\Gamma$. In this sense, the shift to \emph{single letter output alphabet} is
really about enforcing commutativity on the output. However, even with the
restriction to single letter output alphabets, the decidability of
\kl{star-free polyregular functions} among \kl{polyregular functions} remains
an open problem \cite[Conjecture 7.61]{DOUE23}.

\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
as having outputs in $\Nat$, and this is why they were introduced respectively
as \kl{$\Nat$-polyregular functions} ($\NPoly$) and \kl{star-free
$\Nat$-polyregular functions} ($\NSF$) \cite{DOUE21,DOUE22}. From this
point of view, $\NPoly$ is a subclass of a larger and well-studied class of
functions called (non-commutative) \intro{rational series}
\cite{REUT80,BERE88,BERE10}. It has been noticed in \cite{CDTL23} that
\kl{$\Nat$-polyregular functions} correspond to a natural subclass of
\intro{$\Nat$-rational series} ($\intro*\NRat$) that was frequently
rediscovered in the literature \cite{SCHU62,KRRC13}. In \cite{CDTL23}, the
authors introduced the class of \kl{$\Rel$-polyregular functions}
($\ZPoly$) that generalize \kl{$\Nat$-polyregular functions} by allowing
negative outputs. This remains a subclass of \kl{rational series}, more
precisely, of the class of \intro{$\Rel$-rational series} ($\intro*\ZRat$). For
this class of functions the aperiodic variant $\ZSF$ has been shown to
be decidable \cite{CDTL23}. Beware that several non-equivalent notions of
aperiodicity coexist for rational series \cite{REUT80,DRGA19,CDTL23}, and that
we only refer to the one applying to rational series of polynomial growths
introduced in \cite{CDTL23}.

@@

\paragraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
functions} that are furthermore \kl{commutative}. Remark that this enforces a
commutativity property both on the input alphabet and on the output alphabet.
In this restricted setting, we answer positively to \cite[Conjecture
7.61]{DOUE23} (restated in \cref{zsf-nsf:conjecture}), and prove that the
membership problems between $\NSF$, $\ZSF$, $\NPoly$, and $\ZPoly$ are all
decidable, with effective conversion algorithms. This effectively shows that
all arrows in the upper left square of \cref{previously-known-inclusions:fig}
are decidable in the \kl{commutative} setting. Furthermore, we characterize
\kl{commutative} \kl{$\Nat$-polyregular function} as the class of
\kl{$\Rel$-polyregular functions} that are \kl{combinatorial} (defined page
\kpageref{combinatorial}), which is a semantic property akin to the one
\kl{ultimate polynomiality} that characterizes $\ZSF$ inside $\ZPoly$
\cite{CDTL23}. As a byproduct of our analysis, we provide a counter example to
the characterization of polynomials computed by \kl{$\Nat$-rational series}
given by \cite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.

>>>>

\AP There have been recurrent attempts to generalize the notions of regularity
and aperiodicity from languages to functions, leading to classes of various
expressiveness, such as\footnote{This list is far from exhaustive.} Mealy
Machines ($\intro*\Mealy$) \cite{MEAL55}, sequential functions
($\intro*\Sequential$) \cite{SCHU77}, rational functions ($\intro*\Rational$)
\cite{EILE74}, \intro{regular functions} ($\intro*\Regular$) \cite{ENMA02}, and
\intro{polyregular functions} ($\intro*\Poly$) \cite{BOKL19}. For all of these
models, an \emph{aperiodic} counterpart can be defined (respectively,
$\intro*\AMealy$, $\intro*\ASequential$, $\intro*\ARational$,
$\intro*\ARegular$, $\intro*\SF$), lifting the correspondence between
counter-free automata, star-free regular languages and aperiodic monoids to the
functional setting \textbf{TODO: cite more carefully here}
\cite{FKT14,BOJA14,CADA15,DJR16,BDK18,BOKL19,DGK21}. The inclusions between
these classes of functions are all known to be strict, and we depicted in
\cref{computational-models:fig} the status of the \emph{membership problem}
associated to these strict inclusions (that is, given a function $f$, decide if
$f$ can be computed by a function of a proper subclass). Note that in
\cref{computational-models:fig}, if $\mathsf{B}$ is a proper subclass of
$\mathsf{C}$, then the \emph{aperiodic} variant of $\mathsf{B}$ is precisely
the intersection of $\mathsf{B}$ with the \emph{aperiodic} variant of the
larger class $\mathsf{C}$. Because such a (strong) statement does not
explicitly appear in the literature, a standalone proof will be given in
\cref{orthogonal-aperiodic-power:thm}. As an example, to conclude that a
function can be computed by an \emph{aperiodic sequential function}
($\ASequential$) it is enough to prove that it is sequential ($\Sequential$),
and computed by an \intro{aperiodic polyregular function} ($\SF$).

\AP To approach the \emph{aperiodicity} membership problems the traditional
approach has been to develop so-called canonical objects, akin to the minimal
automaton or the syntactic monoid of regular languages, over which aperiodicity
is equivalent to a simple syntactic criterion, e.g. having no counters
\cite{MNPA71}. This is the proof scheme that has been implemented for
sequential functions \cite{CHOF03}, and more recently for rational functions
\cite{FGL16,FGLM18}. Unfortunately, such a canonical object remains to be
defined for regular and polyregular functions.

@@

\AP In order to approach the decision problems on \kl{polyregular functions},
restricting the output alphabet to a single letter has proven to be a fruitful
approach \cite{DOUE21,DOUE22}. The key advantage of single-letter output
alphabets is that the particular order in which letters are produced by the
machine is irrelevant. This restriction greatly simplifies the definition of
\kl{polyregular functions}, and opens the door to results that would otherwise
fail because of this ordering, such as the so-called \emph{pebble minimization
theorem} of \cite{DOUE21}, which fails in the general case
\cite{BOJA22,KLEP23}. Let us remark that the restriction to \emph{unary output}
in the case of \kl{polyregular functions} can often be lifted to outputs in the
free commutative semigroup generated by a finite alphabet $\Gamma$. In this
sense, the shift to \emph{single letter output alphabet} is really about
enforcing commutativity on the output. However, even with the restriction to
single letter output alphabets, the decidability of \kl{star-free polyregular
functions} among \kl{polyregular functions} remains an open problem
\cite[Conjecture 7.61]{DOUE23}.

\AP \kl{Polyregular functions} with a single letter output alphabet can be seen
as having outputs in $\Nat$, and this is why they were introduced respectively
as \kl{$\Nat$-polyregular functions} ($\NPoly$, for \kl{polyregular functions}
with unary output) and \kl{star-free $\Nat$-polyregular functions} ($\NSF$
\kl{for star-free polyregular functions} with unary output)
\cite{DOUE21,DOUE22}. From this point of view, $\NPoly$ is a subclass of a
larger and well-studied class of functions called (non-commutative)
\intro{rational series} \cite{REUT80,BERE88,BERE10}. It has been noticed in
\cite{CDTL23} that \kl{$\Nat$-polyregular functions} correspond to a natural
subclass of \intro{$\Nat$-rational series} ($\intro*\NRat$), another model of
computation that aims at generalising the notion of regular language to
functions with outputs in a semiring \cite{BERE88,BERE10}. It was frequently
rediscovered in the literature that \kl{$\Nat$-polyregular functions} are a
subclass of \kl{rational series} \cite{SCHU62,KRRC13,CDTL23}.

In \cite{CDTL23}, the authors introduced the class of \kl{$\Rel$-polyregular
functions} ($\ZPoly$) that generalize \kl{$\Nat$-polyregular functions} by
allowing negative outputs. This remains a subclass of \kl{rational series},
more precisely, of the class of \intro{$\Rel$-rational series}
($\intro*\ZRat$). For this class of functions the aperiodic variant $\ZSF$ has
been shown to be decidable \cite{CDTL23}. Beware that several non-equivalent
notions of aperiodicity coexist for rational series
\cite{REUT80,DRGA19,CDTL23}, and that we only refer to the one applying to
\kl{rational series} that are actually in $\ZPoly$ \cite{CDTL23}.

@@

\paragraph*{Contributions.} This paper focuses on \kl{$\Nat$-polyregular
functions} that are furthermore \kl{commutative}, that is where the output of
the functions remains unchanged when the letters of the input word are permuted.
In this restricted setting, we answer positively to \cite[Conjecture
7.61]{DOUE23} (restated in \cref{zsf-nsf:conjecture}), and prove that the
membership problems between $\NSF$, $\ZSF$, $\NPoly$, and $\ZPoly$ are all
decidable, with effective conversion algorithms. This effectively shows that
all arrows in the upper left square of \cref{previously-known-inclusions:fig}
are decidable in the \kl{commutative} setting. Furthermore, we characterize
\kl{commutative} \kl{$\Nat$-polyregular functions} as the class of
\kl{$\Rel$-polyregular functions} that are \kl{combinatorial} (defined page
\kpageref{combinatorial}), which is a semantic property akin to the notion of
\kl{ultimate polynomiality} that characterizes $\ZSF$ inside $\ZPoly$
\cite{CDTL23}. As a byproduct of our analysis, we provide a counter example to
the characterization of polynomials computed by \kl{$\Nat$-rational series}
given by \cite[Theorem 3.3, page 4; re-introduced as \cref{karh:thm}]{KARH77}.
